\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage[margin=1in]{geometry}

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{GroundClock: Measuring Whether Language Models Use the Current Time They Are Given}
\author{Thamilvendhan Munirathinam\\ Independent Researcher\\
\texttt{mthamil107@gmail.com} \quad \url{https://github.com/mthamil107}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
Large language models are stateless and frozen at training time: they have no clock and cannot
know the current date. In deployment, applications supply the date by injecting it into the
system prompt on every request, and some additionally expose a clock as a callable tool. A growing
body of work shows that this injection is necessary but not sufficient---even when the current time
is present in context, models frequently fail to \emph{use} it. We introduce \textbf{GroundClock},
a small, provider-agnostic reference layer that supplies a consistent, timezone-aware,
provenance-tagged notion of ``now'' through both a system-prompt block and a \texttt{get\_current\_time}
tool drawn from a single clock, and \textbf{NowBench}, a benchmark whose every gold item has an
answer derivable \emph{only} from the injected instant---so a correct answer is direct evidence the
model used the clock rather than its training-era prior. NowBench scores \emph{alignment} (did it
use the injected time), \emph{usage rate}, cross-turn \emph{consistency}, and, via deliberate
no-match probes, \emph{calibration}. Across four OpenAI models spanning two generations, injecting
the time raises alignment sharply, yet using it correctly for timezone reasoning remains a failure
that is \emph{non-monotonic} in model tier (a small model beats two larger ones), and calibration is
\emph{uniformly absent}---no model abstains when given no clock; a callable clock tool nearly closes
the gap. The contribution is not a new algorithm---date injection, tool
use, and timezone arithmetic are all standard---it is the pairing of a now-provisioning layer with a
benchmark that measures whether the injected time is actually \emph{used}, which prior work has not
shipped jointly.
\end{abstract}

\section{Introduction}
A recurring confusion among practitioners is that a language model ``secretly tracks'' the date. It
does not. The model is a fixed function of its weights; each API call is independent and carries no
clock or cross-call memory. The appearance of date awareness comes entirely from the application
layer: products such as Claude and ChatGPT inject the current date, time, and timezone into the
system prompt on every request \cite{anthropic_sysprompt,gpt55leak,willison_claude4}, holding the
model's training \emph{knowledge cutoff} as a separate, fixed value. The research question is thus
not ``does the model know the date'' but \textbf{``does the model use the date it was given?''}

This distinction matters because the second question has an uncomfortable answer. Recent studies
report that models often ignore injected temporal information \cite{span2025,blindness2025}: across
agentic traces, timestamps appear in a small fraction of reasoning steps, and time-conditioned
behaviors improve only modestly when the time is injected rather than inferred. We call this
\textbf{temporal blindness}: the gap between \emph{having} the current time and \emph{using} it.

\paragraph{Contributions.}
\textbf{C1}~A framing separating the model's absent clock, its fixed cutoff, and per-request time
injection, identifying \emph{use of injected time} as the property of interest.
\textbf{C2}~\textbf{GroundClock}, a reference layer exposing one clock through a system-prompt block
and a \texttt{get\_current\_time} tool, with a structured, timezone-aware, provenance-tagged context
object and a published JSON Schema.
\textbf{C3}~\textbf{NowBench}, five task families in which every gold item is answerable only from
the injected instant.
\textbf{C4}~A \emph{calibration} protocol: no-match probes scored separately so the headline is
honest.
\textbf{C5}~A reproducible harness with a frozen-clock mode, a deterministic offline mock, and a
runnable Claude (\texttt{claude-opus-4-8}) backend.

\section{Background}
\textbf{Models are stateless and frozen.} Weights do not change after training; a call retains
nothing from a prior call, so the model has no inherent ``now''. \textbf{Knowledge cutoff is not a
clock,} and reported cutoffs are unreliable---effective cutoffs vary by topic
\cite{cheng2024dated,shah2025cutoff}. \textbf{Date is supplied by injection} \cite{anthropic_sysprompt,gpt55leak}.
\textbf{Temporal reasoning is hard}: TimeQA \cite{chen2021timeqa}, TempReason
\cite{tan2023tempreason}, and cross-calendar reasoning
\cite{span2025} report large gaps to humans, but measure \emph{knowledge}, not \emph{use} of an
injected now.

\section{GroundClock and NowBench}
\paragraph{Temporal context object.} Fields: \texttt{instant} (UTC, RFC 3339), \texttt{timezone}
(IANA), \texttt{local\_time}, \texttt{utc\_offset} (DST-correct), \texttt{weekday}, \texttt{sequence}
(monotonic per session), \texttt{provenance}, and optional \texttt{knowledge\_cutoff}. Invariants:
(i)~\texttt{local\_time} converts to \texttt{instant}; (ii)~\texttt{sequence} is non-decreasing, so
drift is detectable.

\paragraph{Two surfaces, one clock.} The object renders to a deterministic
\texttt{<temporal\_context>} block and backs a \texttt{get\_current\_time} tool from the same
\texttt{Clock}, so they never disagree; a \texttt{FrozenClock} makes runs reproducible.

\paragraph{Families and conditions.} \emph{present\_date}, \emph{timezone\_now}, \emph{relative\_date},
\emph{deadline}, \emph{consistency}. Conditions: \emph{baseline} (no time), \emph{grounded} (block),
\emph{tool} (block + tool). \textbf{Metrics:} alignment (correct/gold), usage rate, consistency, and
\textbf{calibration} on no-match probes---scored separately and never folded into alignment.

\section{Evaluation}
\paragraph{Offline harness.} A deterministic mock parameterized by a \emph{compliance} rate (the
fraction of items on which it uses an available clock, abstaining when none is present) validates
the plumbing. At \texttt{compliance}~$=1.0$:

\begin{center}
\begin{tabular}{lrrrrr}
\toprule
condition & alignment & usage & consistency & calibration & tool calls \\
\midrule
baseline & 0\% & 0\% & 0\% & 100\% & 0 \\
grounded & 100\% & 100\% & 100\% & 100\% & 0 \\
tool & 100\% & 100\% & 100\% & 100\% & 24 \\
\bottomrule
\end{tabular}
\end{center}

\noindent (20 gold items; calibration on 3 no-match probes, excluded from alignment.) Lower
compliance reproduces partial blindness. The scientific result is the \emph{real} run.

\paragraph{Real run: four OpenAI models.} We evaluate four models across two generations and a
range of sizes under the three conditions; the knowledge-cutoff field is omitted to avoid asserting
an inaccurate value. We report alignment by condition plus two diagnostics: grounded
\texttt{timezone\_now} alignment and probe calibration.

\begin{center}
\begin{tabular}{lrrrrr}
\toprule
model & baseline & grounded & tool & tz\_now (grnd) & calibration \\
\midrule
gpt-4o-mini & 0.10 & 0.95 & 1.00 & 0.75 & 0.00 \\
gpt-4.1 & 0.10 & 0.85 & 0.90 & 0.25 & 0.00 \\
gpt-5.1 & 0.15 & 0.85 & 1.00 & 0.25 & 0.00 \\
gpt-5.5 & 0.15 & 1.00 & 1.00 & 1.00 & 0.00 \\
\bottomrule
\end{tabular}
\end{center}

\noindent (20 gold items; calibration on 3 no-match probes; grounded consistency is $1.00$ for all
four.) \textbf{Findings:} (1)~no model knows ``now'' (baseline $0.10$--$0.15$), and injection lifts
all of them (grounded $0.85$--$1.00$); (2)~the residual failure is timezone reasoning and it is
\emph{non-monotonic in tier}---\texttt{timezone\_now} ranges $0.25$--$1.00$, and the smallest model
(gpt-4o-mini, $0.75$) beats two larger/newer ones; (3)~the \texttt{get\_current\_time} tool nearly
closes the gap (tool $1.00$ for three of four; $0.90$ for gpt-4.1) \cite{span2025}; (4)~\textbf{
calibration is $0$ for every model}---on no-clock probes none ever abstains, each confabulating a
date in all cases. \textbf{Caveats:} four models, small suite; we cannot fully exclude provider-side
date injection, but failing frozen-date items and confabulating probes are consistent with none.
Reproduce: \texttt{nowbench run -{}-provider openai -{}-model <id>}.

\section{Threat Model (condensed)}
The authoritative time source is the host \texttt{Clock}; anything read from the conversation is
untrusted as a time source. Key threats: time spoofing via prompt injection (CWE-345), mitigated by
\texttt{provenance} and the consistency metric; timezone confusion (CWE-697), mitigated by the
local$=$UTC invariant and DST-correct offsets; stale/replayed now (CWE-294 analog), mitigated by the
monotonic \texttt{sequence}. Clock-source compromise is delegated to host hardening.

\section{Related Work}
Temporal-knowledge benchmarks \cite{chen2021timeqa,tan2023tempreason,span2025} measure
what models know; ``Dated Data'' \cite{cheng2024dated} and domain-cutoff studies
\cite{shah2025cutoff} show reported cutoffs are unreliable; the temporal-blindness study
\cite{blindness2025} shows injected time is under-used.
GroundClock is complementary: it assumes injection and tools exist and supplies the missing
piece---a benchmark scoring whether the injected time is \emph{used}, plus a reference layer.

\section{Discussion and Future Work}
NowBench v0 is small and deterministic. Planned: adversarial ``fake-now'' probes, timestamp-usage
detection in reasoning traces, multi-turn agent traces, and additional providers for a cross-model
leaderboard. The durable artifact is the \emph{protocol}---usage and calibration reported
together---not any single number.

\section{Conclusion}
The premise that a model ``tracks'' the date misreads a real phenomenon: the date is supplied
externally, every request, and is often not used. GroundClock supplies it consistently through one
clock and two surfaces; NowBench measures whether it is used, honestly. The contribution is not a
new algorithm but a benchmark-and-reference pairing that turns temporal blindness into a number you
can track.

\section*{Reproducibility}
Code, spec, JSON Schema, threat model, and the full harness (offline mock plus OpenAI and Anthropic
backends) are at \url{https://github.com/mthamil107/groundclock}. \texttt{pytest},
\texttt{scripts/verify\_spec.py}, and \texttt{scripts/run\_microbench.py} reproduce the offline
results; \texttt{nowbench run -{}-provider openai -{}-model <id>} reproduces the model runs.

\bibliographystyle{plain}
\bibliography{groundclock}

\end{document}
